#Reproducible Research Course Assignment:
#Peer Assessment1


##Loading and preprocessing the data

Download the data set from the web source and unzip it at the local "./data" directory
```{r download, echo=TRUE}
if(!file.exists("./data")) {
    dir.create("./data")
}
  
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if(!file.exists("./data/activity")) {
    download.file(fileUrl, destfile="./data/activity.zip", method = "libcurl")
    unzip(zipfile="./data/activity.zip", exdir="./data")
}
```

Loading and preprocessing the data

Convert "Factor" variable to variables of class "Date" representing calendar dates
```{r prep, echo=TRUE}
move_data <- read.csv("./data/activity.csv")
str(move_data)
head(move_data)
tail(move_data)

move_data$date <- as.Date(move_data$date, format = "%Y-%m-%d")
str(move_data)
str(move_data$date)
```

Load R packages for data processing and plotting
```{r packload, echo=TRUE}
library(plyr)
library(ggplot2)
```

##What is mean total number of steps taken per day?

Calculate the total number of steps taken per day
```{r totalsteps, echo=TRUE}
total.steps <- ddply(move_data, .(date), summarize, total_steps = sum(steps))
head(total.steps)
```

Make a histogram of the total number of steps taken each day
```{r totalplot, echo=TRUE}
total.plot <- ggplot(total.steps, aes(x = total_steps))
total.plot <- total.plot + geom_histogram(color = "darkgreen", fill = "white")
print(total.plot)
```

Calculate and report the mean and median of the total number of steps taken per day
```{r totalsummary, echo=TRUE}
summary(total.steps$total_steps)
summary(total.steps$total_steps)["Mean"]
summary(total.steps$total_steps)["Median"]
```

##What is the average daily activity pattern?

Calculate the average number of steps taken per each interval
```{r avesteps, echo=TRUE}
average.steps <- ddply(move_data, .(interval), summarize, average_steps = mean(steps, na.rm = TRUE))
head(average.steps)
tail(average.steps)
```

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r aveplot, echo=TRUE}
average.plot <- ggplot(average.steps, aes(x = interval, y = average_steps))
average.plot <- average.plot + geom_line(color = "red", width = 0.4)
print(average.plot)
summary(average.steps$average_steps)
summary(average.steps$average_steps)["Max."]
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r maxinterval, echo=TRUE}
average.steps[which.max(average.steps$average_steps),]
average.steps[which.max(average.steps$average_steps),]$interval
```

So, the 5-minute interval containing the maximum number of steps is **835** 

##Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r totalnas, echo=TRUE}
table(complete.cases(move_data))
```

Create a new dataset that is equal to the original dataset but with the missing data filled in

Strategy for imputing missing data: Fill in all of the missing values in the dataset using the mean for that 5-minute interval
```{r fillindata, echo=TRUE}
imputed_data <- move_data
imputed_data$steps <- ifelse(is.na(imputed_data$steps), average.steps$average_steps[match(imputed_data$interval, average.steps$interval)], imputed_data$steps)
table(complete.cases(move_data))
table(complete.cases(imputed_data))
```

Make a histogram of the total number of steps taken each day 
```{r totalfillplot, echo=TRUE}
total.imputed.steps <- ddply(imputed_data, .(date), summarize, total_steps = sum(steps))
head(total.imputed.steps)

total.imputed.plot <- ggplot(total.imputed.steps, aes(x = total_steps))
total.imputed.plot <- total.imputed.plot + geom_histogram(color = "dark green", fill = "white")
print(total.imputed.plot)
```

Calculate and report the mean and median total number of steps taken per day
```{r totalfillsummary, echo=TRUE}
summary(total.imputed.steps$total_steps)
summary(total.imputed.steps$total_steps)["Mean"]
summary(total.imputed.steps$total_steps)["Median"]
```

The estimates from the original dataset which ignores the missing values actually are not very different from the mean and median of new dataset imputing missing data.

The histograms of total number of steps from both datasets show a nearly approximated Gaussian distribution.

##Are there differences in activity patterns between weekdays and weekends?

Set time locale as "English_United States.1252" to get the day of week in English
```{r setlocale, echo=TRUE}
Sys.setlocale("LC_TIME", "English")
```

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
```{r weekdiv, echo=TRUE}
move_data$day <- weekdays(move_data$date)
day.week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
move_data$weekdivide <- ifelse(move_data$day %in% day.week, "weekday", "weekend")
head(move_data)
table(move_data$weekdivide)
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r avedayplot, echo=TRUE}
average.day.steps <- ddply(move_data, .(interval, weekdivide), summarize, average_steps = mean(steps, na.rm = TRUE))
head(average.day.steps)

average.day.plot <- ggplot(average.day.steps, aes(x = interval, y = average_steps))
average.day.plot <- average.day.plot + geom_line(color = "red", width = 0.4)
average.day.plot <- average.day.plot + facet_grid(. ~ weekdivide)
print(average.day.plot)
```